Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-10-20 ms/main.tex
Add partial derivatives
revised external stability primer

\subsection*{Modifier model}

@@

Therefore, if a modifier allele that increases social association invades the population, the population mean fitness will increase. But can such a modifier allele invade the population?
We assume that social association is determined by a genetically transmitted locus with alleles $m$ and $M$, which induce association probabilities $\alpha_1$ and $\alpha_2$, respectively.

We define the frequencies of the pheno-genotypes $Am$, $Bm$, $AM$, and $BM$ as $\hat{p}_1$, $\hat{p}_2$, $\hat{p}_3$, and $\hat{p}_4$, respectively. 
Then the following recursion system defines the frequencies of the pheno-genotypes in the next generation,
\begin{equation} \label{eq:next_gen_p_1}
  \begin{aligned}
	% p1
  \bar{w}\tilde{p}_1' & = \tilde{p}_1(\tilde{p}_1+\hat{p_3})(1+b-c)(1 - (1-\alpha_1)(\tilde{p}_2 + \tilde{p}_4)T_B) \\
  &+ \tilde{p}_1(\tilde{p}_2 + \tilde{p}_4)(1-c)(1-\alpha_1T_B(\hat{p_1}+\tilde{p}_3) - T_B(\tilde{p}_2+\tilde{p}_4)) \\
  &+ \tilde{p}_2(\tilde{p}_1 + \tilde{p}_3)(1+b)T_A(\tilde{p}_1 + \tilde{p}_3 + \alpha_1(\tilde{p}_2 + \hat{p_4})) \\
  &+ \tilde{p}_2(\tilde{p}_2 + \tilde{p}_4)(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_1)T_A 
\\
	% p2
  \bar{w}\tilde{p}_2' & = \tilde{p}_2(\tilde{p}_2+\hat{p_4})(1 - (1-\alpha_1)(\tilde{p}_1 + \tilde{p}_3)T_A) \\
  &+ \tilde{p}_2(\tilde{p}_1 + \tilde{p}_3)(1+b)(1-\alpha_1T_A(\tilde{p}_2 + \tilde{p}_4)-T_A(\tilde{p}_1 + \tilde{p}_3))\\
  &+ \tilde{p}_1(\tilde{p}_2 + \tilde{p}_4)(1-c)(\alpha_1T_B + (1-\alpha_1)(\tilde{p}_2+\tilde{p}_4)T_B) \\
  &+ \tilde{p}_1(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_1)(\tilde{p}_2+\tilde{p}_4)T_B 
\\
	% p3
  \bar{w}\tilde{p}_3' & = \tilde{p}_3(\tilde{p}_1+\hat{p_3})(1+b-c)(1 - (1-\alpha_2)(\tilde{p}_2 + \tilde{p}_4)T_B) \\
  &+ \tilde{p}_3(\tilde{p}_2 + \tilde{p}_4)(1-c)(1-\alpha_2T_B(\hat{p_1}+\tilde{p}_3) - T_B(\tilde{p}_2+\tilde{p}_4)) \\
  &+ \tilde{p}_4(\tilde{p}_1 + \tilde{p}_3)(1+b)T_A(\tilde{p}_1 + \tilde{p}_3 + \alpha_2(\tilde{p}_2 + \hat{p_4})) \\
  &+ \tilde{p}_4(\tilde{p}_2 + \tilde{p}_4)(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_2)T_A 
\\
	% p4
  \bar{w}\tilde{p}_4' & = \tilde{p}_4(\tilde{p}_2+\hat{p_4})(1 - (1-\alpha_2)(\tilde{p}_1 + \tilde{p}_3)T_A) \\
  &+ \tilde{p}_4(\tilde{p}_1 + \tilde{p}_3)(1+b)(1-\alpha_2T_A(\tilde{p}_2 + \tilde{p}_4)-T_A(\tilde{p}_1 + \tilde{p}_3))\\
  &+ \tilde{p}_3(\tilde{p}_2 + \tilde{p}_4)(1-c)(\alpha_2T_B + (1-\alpha_2)(\tilde{p}_2+\tilde{p}_4)T_B) \\
  &+ \tilde{p}_3(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_2)(\tilde{p}_2+\tilde{p}_4)T_B \;,
  \end{aligned}
\end{equation}
where $\bar{w} = 1 + (b-c)\cdot(\tilde{p}_1 + \tilde{p}_3)$ is the population mean fitness.

Denote this system by $\bar{w} \vec{\tilde{p}} = f(\vec{\tilde{p}})$, where $\vec{\tilde{p}} = (\tilde{p}_1,\tilde{p}_2,\tilde{p}_3,\tilde{p}_4)$.
We are interested in the invasion of allele $M$ to a population at equilibrium with allele $m$, that is, with equilibrium frequencies $\vec{\tilde{p}}^*=(\tilde{p}^*, 1-\tilde{p}^*, 0, 0)$, with $\tilde{p}^*$ defined in
\autoref{eq:vert_hori_equilibrium}.
The success of the invasion of allele $M$ is determined by the external stability matrix $\mathbf{L}^*_{ex}$~\citep{Liberman1988}, which is derived from \autoref{eq:next_gen_p_1} as
\begin{equation}
\mathbf{L}^*_{ex} = \frac{1}{\bar{w}^*} \begin{bmatrix}
	 \frac{\partial \tilde{p}_3'}{\partial \tilde{p}_3}(\vec{\tilde{p}}^*) &
	 \frac{\partial \tilde{p}_3'}{\partial \tilde{p}_4}(\vec{\tilde{p}}^*) \\
	 \frac{\partial \tilde{p}_4'}{\partial \tilde{p}_3}(\vec{\tilde{p}}^*) &
	 \frac{\partial \tilde{p}_4'}{\partial \tilde{p}_4} (\vec{\tilde{p}}^*)
\end{bmatrix} \;.
\end{equation}
If the leading eigenvalue of $\mathbf{L}^*_{ex}$ is less than 1, then the equilibrium $\vec{\tilde{p}}$ is locally stable and allele $M$ cannot invade the population of allele $m$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{With Population Structure}

>>>>

\subsection*{Modifiers of social association}

@@

Therefore, if the social association increases, the population mean fitness will increase.
But can an allele that increases the social association evolve? 
To answer this question, we extend our model to include a ``modifier locus''~\citep{Liberman1986,Liberman1988}.
The function of this modifier locus is to determine the social association.
It has no direct effect on fitness.
We assume that the modifier locus has two alleles, $M$ and $m$, which induce association probabilities $\alpha_1$ and $\alpha_2$, respectively, and that initially only $M$ is present.
Suppose that the population evolves to a stable equilibrium when only $M$ is present; we check the local stability of this equilibrium to the introduction of the allele $m$ at the modifier locus. This is called ``external stability''; For more details on external stability, see \citet{Altenberg2017}.

We define the frequencies of the pheno-genotypes $AM$, $BM$, $Am$, and $Bm$ as $\vec{\tilde{p}}= (\tilde{p}_1$, $\tilde{p}_2$, $\tilde{p}_3$, $\tilde{p}_4)$. 
Then the following recursion system defines the frequencies of the pheno-genotypes in the next generation,
\begin{equation} \label{eq:next_gen_p_1}
  \begin{aligned}
	% p1
  \bar{w}\tilde{p}_1' & = \tilde{p}_1(\tilde{p}_1+\tilde{p}_3)(1+b-c)(1 - (1-\alpha_1)(\tilde{p}_2 + \tilde{p}_4)T_B) \\
  &+ \tilde{p}_1(\tilde{p}_2 + \tilde{p}_4)(1-c)(1-\alpha_1T_B(\tilde{p}_1 + \tilde{p}_3) - T_B(\tilde{p}_2 + \tilde{p}_4)) \\
  &+ \tilde{p}_2(\tilde{p}_1 + \tilde{p}_3)(1+b)T_A(\tilde{p}_1 + \tilde{p}_3 + \alpha_1(\tilde{p}_2 + \tilde{p}_4)) \\
  &+ \tilde{p}_2(\tilde{p}_2 + \tilde{p}_4)(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_1)T_A 
\\
	% p2
  \bar{w}\tilde{p}_2' & = \tilde{p}_2(\tilde{p}_2+\tilde{p}_4)(1 - (1-\alpha_1)(\tilde{p}_1 + \tilde{p}_3)T_A) \\
  &+ \tilde{p}_2(\tilde{p}_1 + \tilde{p}_3)(1+b)(1-\alpha_1 T_A(\tilde{p}_2 + \tilde{p}_4) - T_A(\tilde{p}_1 + \tilde{p}_3))\\
  &+ \tilde{p}_1(\tilde{p}_2 + \tilde{p}_4)(1-c)(\alpha_1 T_B + (1-\alpha_1)(\tilde{p}_2 + \tilde{p}_4)T_B) \\
  &+ \tilde{p}_1(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_1)(\tilde{p}_2 + \tilde{p}_4)T_B 
\\
	% p3
  \bar{w}\tilde{p}_3' & = \tilde{p}_3(\tilde{p}_1+\tilde{p}_3)(1+b-c)(1 - (1-\alpha_2)(\tilde{p}_2 + \tilde{p}_4)T_B) \\
  &+ \tilde{p}_3(\tilde{p}_2 + \tilde{p}_4)(1-c)(1-\alpha_2T_B(\tilde{p}_1 + \tilde{p}_3) - T_B(\tilde{p}_2 + \tilde{p}_4)) \\
  &+ \tilde{p}_4(\tilde{p}_1 + \tilde{p}_3)(1+b)T_A(\tilde{p}_1 + \tilde{p}_3 + \alpha_2(\tilde{p}_2 + \tilde{p}_4)) \\
  &+ \tilde{p}_4(\tilde{p}_2 + \tilde{p}_4)(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_2)T_A 
\\
	% p4
  \bar{w}\tilde{p}_4' & = \tilde{p}_4(\tilde{p}_2 + \tilde{p}_4)(1 - (1-\alpha_2)(\tilde{p}_1 + \tilde{p}_3)T_A) \\
  &+ \tilde{p}_4(\tilde{p}_1 + \tilde{p}_3)(1+b)(1-\alpha_2T_A(\tilde{p}_2 + \tilde{p}_4)-T_A(\tilde{p}_1 + \tilde{p}_3))\\
  &+ \tilde{p}_3(\tilde{p}_2 + \tilde{p}_4)(1-c)(\alpha_2T_B + (1-\alpha_2)(\tilde{p}_2 + \tilde{p}_4)T_B) \\
  &+ \tilde{p}_3(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_2)(\tilde{p}_2 + \tilde{p}_4)T_B \;,
  \end{aligned}
\end{equation}
%which we can write as 
%\begin{equation}
%\bar{w} \vec{\tilde{p}} = F(\vec{\tilde{p}}),
%\end{equation}
where $\bar{w} = 1 + (b-c)\cdot(\tilde{p}_1 + \tilde{p}_3)$ is the population mean fitness.% and $\vec{\tilde{p}} = (\tilde{p}_1,\tilde{p}_2,\tilde{p}_3,\tilde{p}_4)$.

The local stability of $\vec{\tilde{p}^*} = (\tilde{p}^*, 1-\tilde{p}^*, 0, 0)$, the equilibrium when only allele $M$ is present (see \autoref{eq:vert_hori_equilibrium}), to the introduction of allele $m$ is determined by the linear approximation $\vec{L}^*$ of the transformation in \autoref{eq:next_gen_p_1} near $\vec{\tilde{p}^*}$ (i.e., the Jacobian of the transformation at the equilibrium).
$\vec{L}^*$ is known to have a block structure, with the diagonal blocks occupied by the matrices $\vec{L}^*_{in}$ and $\vec{L}^*_{ex}$.
The latter is the external stability matrix: the linear approximation to the transformation near $\vec{\tilde{p}^*}$ involving only the phono-genotypes $Am$ and $Bm$. Because we assume that $\vec{\tilde{p}^*}$ is internally stable (i.e. locally stable to small perturbations in the frequencies of $AM$ and $BM$), the stability of 
$\vec{\tilde{p}^*}$ is determined by the eigenvalues of the external stability matrix $\vec{L}^*_{ex}$, which is derived from \autoref{eq:next_gen_p_1} as
\begin{equation}
\mathbf{L}^*_{ex} = \frac{1}{\bar{w}^*} \begin{bmatrix}
	 \frac{\partial \tilde{p}_3'}{\partial \tilde{p}_3}(\vec{\tilde{p}}^*) &
	 \frac{\partial \tilde{p}_3'}{\partial \tilde{p}_4}(\vec{\tilde{p}}^*) \\
	 \frac{\partial \tilde{p}_4'}{\partial \tilde{p}_3}(\vec{\tilde{p}}^*) &
	 \frac{\partial \tilde{p}_4'}{\partial \tilde{p}_4} (\vec{\tilde{p}}^*)
\end{bmatrix} \;.
\end{equation}
If the eigenvalues of $\mathbf{L}^*_{ex}$ are less than 1, then the equilibrium $\vec{\tilde{p}}$ is externally stable and allele $m$ cannot invade the population of allele $M$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection*{With Population Structure}
